Simulation Study

### 2020-09-15 simulation_studies/draft_sim_study_13_sep.Rmd
2-dim case sim study done

In order ensure legibility we employ some helper functions for model initialization, plotting, etc. These and the packages are loaded in the follwing chunk.

```{r}
# setting the stage
rm(list=ls())
options(scipen=999)
source("helper_plot_model.R")
source("helper_init_model.R")
source("helper_calc_mse.R")
source("helper_init_large_model.R")
library(asp20boost)K
library(gamboostLSS)
library(microbenchmark)
set.seed(1337)
```

@@

@@

## Small Model - 4 Parameter Case

The goal in this first part is to demonstrate, that the algorithm finds correct estimates for sensible choices of \beta and \gamma. We use the following 5 differing parameter configurations to explore location-scale regression model characteristics.

@@

# config 3 ----------------------
beta_3 <- c(0,7)
gamma_3 <- c(-3,3)

# config 4 ----------------------
beta_4 <- c(0,1)
gamma_4 <- c(3,-3)

# config 5 ----------------------
beta_5 <- c(0,1)
gamma_5 <- c(0.3,-0.3)

# config 6 ----------------------
beta_6 <- c(0,15)
gamma_6 <- c(1,-1)
```

@@

This parameter configuration is equivalent to the example case of the asp20model package. A model object of 1000 observations is created and the model is plotted by employing the helper functions mentioned above.

```{r}
mod_1 <- init_model(beta_1, gamma_1, 1000)$model
plot_model(beta_1, gamma_1, 1000)
```

As seen above the range of response values, as well as the mean increase as the covariate x increases. The black dashed line here, however, does not represent the 95% confidence interval as is usually the case, but rather the standard deviation of the response at the respective covariate value. The lower dashed line represents one standard deviation departing from the expected response in negative direction - the upper dashed line, in turn, in positive direction. Hence the interval of the dashed lines span over two standard deviations.

The red arrow indicates the standard deviation at the point x = 1,  that is the upper limit of covariate values. We designed covariate values to be on the unit interval for convenience. For cases with increasing scale, the red arrow hence indicates maximum standard deviation of the response.

The blue arrow depicts the slope of the location parameter vector *beta_1.*

The ratio of those two quantities is essential for estimability of the model. Here the blue arrow is dominant, hence the model may well be estimated. We will see the opposite case in the following subpart.

```{r}
gradient_boost(mod_1)
round(mod_1$beta, 3)
round(mod_1$gamma, 3)
```

We see that the estimates are close to the true model parameters - for *beta* as well as for *gamma.*

@@

Configuration 2 sets all parameters set to value 2. Estimating this model with the gradient_boost() function does not yield valid results.

```{r}
init <- init_model(beta_2, gamma_2, 1000)
mod_2 <- init$model
gradient_boost(mod_2)
mod_2$beta
mod_2$gamma
```

To understand why this happens we need to understand the model characteristics resulting from parameter configuration 2-2-2-2.

```{r}
plot_model(beta_2, gamma_2, 1000)
```

Here, the red arrow is of large size, while the blue arrow is hardly visible. This is due to the presence of huge standard deviations in response \y, concealing possible changes in expectation of \y. In fact, the standard deviation at x = 1 is:

```{r}
# sd(x = 1) = exp(2 + 2 * 1) = 
exp(gamma_2[1] + gamma_2[2] * 1)
```
The blue arrow indicates change of expected response y due to a one unit increase in the regressor x, which is \beta1 = 2. The estimable signal perishes, due to the high amount of noise caused by the high gamma-parameters. Resulting is a signal-noise ration, that is impossible to estimate. Values for parameters gamma need to be chosen sensibly for simulating data. This is no straight-forward task, since the effect of \gamma on the response standard deviation is not linear, but exponential. To check in the simple case, the model plot containing arrows are of help.

@@

```{r}
mod_3 <- init_model(beta_3, gamma_3, 1000)$model
plot_model(beta_3, gamma_3, 1000)
```

Configuration 3 yields a good signal-noise ratio.

```{r}
gradient_boost(mod_3)
round(mod_3$beta, 3)
round(mod_3$gamma, 3)
```

The algorithm performs well with estimates close to true parameters. In order quantify proximity of estimated model to true data generating closeness we employ the mean squared error. This enables a comparison of model fit for configuration 1 and 3.

```{r}
mse_1 <- calc_MSE(beta_3, gamma_3, mod_3$beta, mod_3$gamma)
mse_1
```

```{r}
mse_3 <- calc_MSE(beta_1, gamma_1, mod_1$beta, mod_1$gamma)
mse_3
```

Model 3 yields a slightly better fit in terms of MSE as model 1, differing only at the 4. decimal place. Realtively, it improved by about 9 percent.

```{r}
round((mse_3$total - mse_1$total) / mse_1$total * 100, digits = 2) #percent
```

@@

```{r}
mod_4 <- init_model(beta_4, gamma_4, 1000)$model
gradient_boost(mod_4)
round(mod_4$beta, 3)
round(mod_4$gamma, 3)
```

Configuration 4, again, is not estimable by the gradient_boost() function. A look at the model plot helps also in this case.

```{r}
plot_model(beta_4, gamma_4, 1000)
```

Here SDs are considerably high for x = 0, namely exp(3) = 20.1, however decreasing fast until reaching a value of exp(3 - 3) = 1 at x = 1. With a beta slope of 1, the algorithm runs into numerical problems yielding NaNs as results.

#### Evaluation Configuration 5

[insert model configuration as formula]

Changing the gamma parameters by factor 0.1 changes model characteristics considerably. Signal-noise ratio is well balanced in this case. The algorithm is able to detect the true parameters considerably close, however not satisfactorily.

```{r}
mod_5 <- init_model(beta_5, gamma_5, 1000)$model
plot_model(beta_5, gamma_5, 1000)
```

```{r}
gradient_boost(mod_5)
round(mod_5$beta, 3)
round(mod_5$gamma, 3)
```

### Evaluation Configuration 6

For robustness, we check a configuration with high location slope and comparably low gamma parameters, resulting in a very good signal-noise ratio. This induces the expectation of good estimates, as seen in configuration 3.

```{r}
mod_6 <- init_model(beta_6, gamma_6, 1000)$model
plot_model(beta_6, gamma_6, 1000)
```

```{r}
gradient_boost(mod_6)
round(mod_6$beta, 3)
round(mod_6$gamma, 3)
```

It turns out that the algorithm performs very poorly. Is does not run into numerical problems, but it yields estimate far away from the true parameter values, leading to high MSE.

```{r}
calc_MSE(beta_6, gamma_6, mod_6$beta, mod_6$gamma)
```

This poor behaviour remains without explanation for now.

@@

The observed 4 parameter cases let us conclude that not every arbitrary parameter configuration is estimable by the implemented gradient_boost() algorithm. A first insight is that signal-noise ratio needs to be substantially high in order to prevent that location effect gets concealed by inflated standard deviations. Second, we observe poor behaviour if negative scale slopes are involved. Since we have no explanation for this behaviour we exclude these in our further assessment.

## Large Model - Variable Selection

@@

For models with many regressors componentwise boosting allows to perform variable selection.

```{r}
# initialize model-----------------------------------------------
init_1 <- init_large_model(beta_range = c(5,10), 
                           gamma_range = c(0.5,1), 
                           dim_param = 5, 
                           int_gamma = -0.2,
                           n = 100)
mod_1 <- init_1$model

# check model via plotting --------------------------------------
# par(mfrow = c(2,2))
# for(i in 1:4) plot(X[,i], y, ylim = c(0,200))

Checking the constructed response is crucial: is the signal-noise ratio appropriate? Unfortunately, this can not be done easily by plotting - as in the 2-dimensional case. In the multidimensional case it helps to look at the n = 100 unit-wise location and scale predictors. Mean and sd of these 100 values help getting an overview.

```{r}
mean(init_1$fitted_response)
sd(init_1$fitted_response)
mean(init_1$fitted_sd)
sd(init_1$fitted_sd)
```

Histograms also may help.

```{r}
par(mfrow=c(2,1))
hist(init_1$fitted_response, breaks = 25)
hist(init_1$fitted_sd, breaks = 25)
```

A balanced signal-noise ratio  is assured, as expected response values span the range of 8 to 22 (varaiteion 14), while response SD cover values of 1 to 4. A variantion amount of 14 is towards a mixumu sd of 4, while way more sds are at about 2.5 and lower. This should result in estimable effect sizes. We have a look at the randomly drawn parameter values for beta and gamma:
```{r}
init_1$beta
init_1$gamma
```

And now boost and find the non-zeros.

```{r}
gradient_boost(mod_1, stepsize = c(0.1,0.01), maxit = 1000, componentwise = TRUE)

mod_1$beta
mod_1$gamma
```

For location, the right parameters are chosen, however not at the proper values. For scale too many parameters become chosen. This may be improved by setting less maxit. However this may lead to move away from the yet quite optimal choice for *beta*. It could also be improved by lowering stepsize for gamma. This results in a pattern where the latter entries of gamma are chosen exactly wrong. We will observe this pattern also in the next examples and have no explanation.

```{r}
mod_1$beta <- rep(0,6)
mod_1$gamma <- rep(0,6)
gradient_boost(mod_1, stepsize = c(0.1,0.001), maxit = 1000, componentwise = TRUE)

mod_1$beta
mod_1$gamma
```

@@

```{r}
init_2 <- init_large_model(c(3,6), c(0.2,0.8), 10, -0.2, n = 100)
mod_2 <- init_2$model
```

```{r}
par(mfrow=c(2,1))
hist(init_2$fitted_response, breaks = 25)
hist(init_2$fitted_sd, breaks = 25)
```

@@

@@

In order to assess our algorithm we compare it to the existing package gamboostLSS. It is the common way in R to adress location-scale regression models by the boosting method.

@@

@@

@@

```{r}
init <- init_model(beta_1, gamma_1, 1000)
mod_1 <- init$model
response_1 <- init$response
covariate_1 <- init$covariate
plot_model(beta_1, gamma_1, 1000)
```

@@

### Evaluation Configuration 3

>>>>

In order ensure legibility we employ some helper functions for model initialization, plotting, etc. These and the packages are loaded in the following chunk.

```{r}
# setting the stage
rm(list=ls())
options(scipen=999)
source("helper_plot_model.R")
source("helper_init_model.R")
source("helper_calc_mse.R")
source("helper_init_large_model.R")
library(asp20boost)
library(gamboostLSS)
library(microbenchmark)
set.seed(1337)
```

@@

Second, we observe poor behaviour if negative scale slopes are involved. Since we have no explanation for this behaviour we exclude these in our further assessment.

@@

## Small Model - Two-dimnesional Case

The goal in this first part is to demonstrate, that the algorithm finds correct estimates for sensible choices of \beta and \gamma. We use the following 4 different parameter configurations to explore location-scale regression model characteristics.

@@

# config 3 ----------------------
beta_3 <- c(0,1)
gamma_3 <- c(-1,-0.3)

# config 4 ----------------------
beta_4 <- c(0,15)
gamma_4 <- c(1,-1)

# config 7 ----------------------
# beta_7 <- c(0,1)
# gamma_7 <- c(3,-3)
```

```{r}
mod_1 <- init_model(beta_1, gamma_1, 1000)$model
gradient_boost(mod_1)
mod_2 <- init_model(beta_2, gamma_2, 1000)$model
gradient_boost(mod_2)
mod_3 <- init_model(beta_3, gamma_3, 1000)$model
gradient_boost(mod_3)
mod_4 <- init_model(beta_4, gamma_4, 1000)$model
gradient_boost(mod_4)
```

```{r, cache=TRUE, fig.height = 8, fig.width = 11}
par(mfrow = c(2,2))
plot_model(beta_1, gamma_1, 1000, "Config 1")
plot_model(beta_2, gamma_2, 1000, "Config 2")
plot_model(beta_3, gamma_3, 1000, "Config 3")
plot_model(beta_4, gamma_4, 1000, "Config 4")
```
```{r}
result_df <- data.frame("Config" = c(1,1,2,2,3,3,4,4),
                        "beta_true" = c(beta_1, beta_2, beta_3, beta_4),
                        "beta est." = c(round(mod_1$beta, 3), round(mod_2$beta, 3), round(mod_3$beta, 3), round(mod_4$beta, 3)),
                        "gamma_true" = c(gamma_1, gamma_2, gamma_3, gamma_4),
                        "gamma est." = c(round(mod_1$gamma, 3), round(mod_2$gamma, 3), round(mod_3$gamma, 3), round(mod_4$gamma, 3))
                        )
result_df
```

@@

This parameter configuration is equivalent to the example case of the *asp20model* package. 
As already mentioned, heteroskedasticity is present and indicated by the funnel shape of the point cloud.
The solid black line represents the true response mean, while the dashed lines represent the true upwards and downwards standard deviations from this mean.
The red arrow indicates the standard deviation at the point x = 1,  that is the upper limit of covariate values, since covariates are design to be on the unit interval.
For cases with increasing scale, the red arrow hence indicates maximum standard deviation of the response.
The blue arrow depicts the slope of the location parameter vector *beta_1.*
The ratio of those two quantities is essential for estimability of the model. 
Here the blue arrow is dominant, hence the model may well be estimated. 
Estimates are displayed in *table[...]* above. We see that both *betahat* and *gammahat* estimates are close to the true parameter vectors *beta* and *gamma.*

@@

Configuration 2 is a somewhat naive parameter configuration, where all parameters are set to value 2. 
Estimating this model does not yield valid results.
The gradient_boost() function runs into numerical problems.
To understand why this happens we need to understand the model characteristics resulting from parameter configuration 2-2-2-2.
Here, the red arrow is of large size, while the blue arrow is hardly visible. 
This is due to the presence of huge standard deviations in response *y*, concealing changes in expectation of *y.* 
In fact, the standard deviation at x = 1 is:

```{r}
exp(gamma_2[1] + gamma_2[2] * 1)
```
In contrast the response expectation values range from 0 to 2.
The blue arrow indicates change of expected response *y* due to a one unit increase in the regressor *x*, which is *beta1* = 2. 
The estimable signal perishes, due to the high amount of noise caused by the high gamma-parameters. 
Resulting is a signal-noise ration, that is impossible to estimate. 
Values for parameters gamma need to be chosen sensibly for simulating data. 
This is no straight-forward task, since the effect of *gamma* on the response standard deviation is not linear, but exponential. 
To check this in the two-dimensional case, the arrows in the model plot are of help.
In the following part handling multidimensional cases, we'll introduce another form of checking.

@@

Configuration 3, in turn, yields reasonable estimates.
It contains a negative gamma slope, leading to decreasing response standard deviations.
This can be observed by the funnel shape pointing in the opposite direction as before.

It may be of interest to check if estimates are fit better (if they lie closer to the true parameters) than e.g. in configuration 1.
To do this we measure the gap between true values and estimates by the mean squared error.
This enables comparison of configurations 1 and 3.

```{r}
mse_1 <- calc_MSE(beta_1, gamma_1, mod_1$beta, mod_1$gamma)
mse_3 <- calc_MSE(beta_3, gamma_3, mod_3$beta, mod_3$gamma)
cbind("config 1" = mse_1,
      "config_3" = mse_3)
```

Configuration 3 leads to smaller MSE, especially for scale estimates *gamma_hat*.
We conclude that our alogrithm performs better in configuration 3.

@@

In configuration 4, we again included a negative gamma slope, as well as a very high beta slope.
The idea is to ensure a very good signal noise ratio, in order to ensure good model fit, indicated by low MSE.
Contrary to our expectations, the algorithm performs poorly for this parameter configuration.

```{r}
mse_4 <- calc_MSE(beta_4, gamma_4, mod_4$beta, mod_4$gamma)
cbind("config 1" = mse_1,
      "config_3" = mse_3,
      "config 4" = mse_4)
```

The bad model fit is due to the fact, that the location slope is exceptionally underestimated.

@@

The observed two-dimensional cases let us conclude that not every arbitrary parameter configuration is estimable by the implemented gradient_boost() algorithm. A first insight is that signal-noise ratio needs to be sufficiently high in order to prevent that location effect gets concealed by inflated standard deviations.

## Large Model - Variable Selection in the multidimensional case

### ToDo

This may be improved by setting less maxit. However this may lead to move away from the yet quite optimal choice for *beta*. It could also be improved by lowering stepsize for gamma. This results in a pattern where the latter entries of gamma are chosen exactly wrong. We will observe this pattern also in the next examples and have no explanation.

@@

For models with many regressors componentwise boosting allows to perform variable selection. The helper function *init_large_model* creates sample data. The parameter vectors beta and gamma are of size *dim_param* and contain some zeros. The non-zero elements are assigned random values in the specified ranges.

```{r}
# initialize model-----------------------------------------------
init_1 <- init_large_model(beta_range = c(5,10), 
                           gamma_range = c(0.5,1), 
                           dim_param = 5, 
                           int_gamma = -0.2,
                           n = 100)
mod_1 <- init_1$model
cbind(beta_true = init_1$beta, gamma_true = init_1$gamma)
```

In *mod_1* beta and gamma ranges are chosen in order to create a model that is estimable. In the multidimensional it is again important to create data with a balanced signal-noise ratio. If gamma parameters add up to high values, the resulting simulated response has huge standard deviation, which conceals the location effects.

In the twodimensional case we used plots to check this ratio. Now, we have a look at the unit-wise location predictors *eta_mu*, as well as the exponentiated scale predictors *exp(eta_sigma)*, which determine the response sd.

```{r}
cbind("eta_mu" = rbind(mean = mean(init_1$fitted_response), var = var(init_1$fitted_response)), 
      "exp_eta_sigma" = rbind(mean = mean(init_1$fitted_sd), var = var(init_1$fitted_sd)))
```

Plotting the empirical distributions of *eta_mu* and *exp_eta_sigma* as histograms gives a visual and more detailed overview of the location-scale regression model characteristics given the assigned parameters *beta* and *gamma*

```{r}
par(mfrow=c(2,1))
hist(init_1$fitted_response, breaks = 25, main = "eta_mu")
hist(init_1$fitted_sd, breaks = 25, main = "exp_eta_sigma")
```

Location predictors take values between 8 and 22, peaking at 15. In contrast, the highest response-SD is 4, while most values are at about 2.5 and lower. The resulting response variation should be small enough to not conceal the location effects, as well as big enough to be estimated.

Performing the boosting alogrithm by calling the gradient_boost() function leads to the follwing results.

```{r}
gradient_boost(mod_1, stepsize = c(0.1,0.001), maxit = 1000, componentwise = TRUE)

cbind(true = init_1$beta, estimate = round(mod_1$beta,2))
cbind(true = init_1$gamma, estimate = round(mod_1$gamma,2))
```

The algorithm depicts correctly which effects on response location exists, that is it finds the non-zero elements of *beta*.
For scale estimate *gamma_hat* too many parameters become chosen, since two zero-elements of true parameter *gamma* are assigned non-zero values by the estimate *gamma_hat*.

The values estimated differ considerably from the true parameter values. 
The resulting MSE is higher than in the four parameter case.

```{r}
calc_MSE(init_1$beta, init_1$gamma, mod_1$beta, mod_1$gamma)
```

@@

```{r}
init_2 <- init_large_model(c(3,6), c(0.2,0.8), 10, -0.2, n = 100)
mod_2 <- init_2$model
cbind(beta_true = init_2$beta, gamma_true = init_2$gamma)
```

```{r}
par(mfrow=c(2,1))
hist(init_2$fitted_response, breaks = 25, main = "eta_mu")
hist(init_2$fitted_sd, breaks = 25, main = "exp_eta_sigma")
```

@@

### Setup 3

@@

In order to assess our algorithm we compare it to the existing package *gamboostLSS.* Therefore we use configurations one and three of the two-dimensional cases

@@

In the *gamboostLSS* package boosting a location-scale regression model works as follows.

```{r}
init <- init_model(beta_1, gamma_1, 1000)
mod_1 <- init$model
response_1 <- init$response
covariate_1 <- init$covariate

@@

### Recap Configuration 1

@@

@@

### Recap Configuration 3
